---
title: "An {arsenal} of R Functions for Statistical Summaries"
subtitle: "https://github.com/mayoverse/arsenal"
author: "Ethan Heinzen, Beth Atkinson, Jason Sinnwell"
date: "2021-08-26"
output: 
  mayodown::mayobeamer:
    slide_level: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

# Motivation

- Mayo Clinic: 3-site research hospital

- Historically a SAS shop

- Now ~500 R users

- A SAS license negotiation revealed our dependency on SAS

- How to port in-house macros and other SAS procedures to R?

# rlocal

- First started with an internal-only package: `rlocal`

- Contained some "private" functions, some "public" functions

- Decided to separate "public" into a CRAN package

# arsenal

- Goal: mimic and improve SAS functionality "easily"

- v0.1.2 on CRAN Dec 30, 2016

- Major releases Feb 2018, Jan 2019, and March 2019

- Minor releases every couple of months

# Functions

- `tableby()` $\equiv$ `%table()`: create a table 1

- `paired()` $\equiv$ `%paired()`: create a table 1 for paired data

- `modelsum()` $\equiv$ `%modelsum()`: fit models over a set of independent variables

- `freqlist()` $\equiv$ `PROC FREQ`: compile frequency table

- `comparedf()` $\equiv$ `PROC COMPARE`: compare two datasets

- `write2*()` $\approx$ `ODS OUTPUT`: output results to a file

# The dataset

\tiny

```{r}
library(arsenal)
data(mockstudy)
str(mockstudy)
```

\normalsize

# tableby

A simple pipe table:

\tiny

```{r}
tb <- tableby(arm ~ sex + age, data = mockstudy)
summary(tb, text = TRUE)
```

\normalsize

# tableby

The markdown equivalent:

\tiny

```{r results='asis'}
tb <- tableby(arm ~ sex + age, data = mockstudy)
summary(tb)
```

Notice that sex (a categorical) is treated differently from age (continuous). `tableby()` supports categoricals (character, logical, factor),
numerics, ordered factors, `survival::Surv()` objects, dates, and results from `arsenal::selectall()`.

\normalsize

# tableby

Common requests:

- Change labels: `set_labels()`, `labels()<-`, `labelTranslations=`

- Change summary statistics: `tableby.control()` or inline

- Change statistical test (p-value): `tableby.control()` or inline, or `modpval.tableby()`.

- Change decimal points: `tableby.control()` or inline


# tableby

\tiny

```{r results='asis'}
tb <- tableby(arm ~ fe(sex, digits.pct = 0) + notest(age, digits = 1, "median", "q1q3"),
              data = mockstudy)
summary(tb, pfootnote = TRUE)
```

\normalsize

# tableby

Without a by-variable:

\tiny

```{r results='asis'}
tb <- tableby( ~ sex + age, data = mockstudy)
summary(tb)
```

\normalsize

# tableby

A stratified, subsetted, and multiple endpoint summary:

\tiny

```{r results='asis'}
tb <- tableby(list(arm, sex) ~ age, strata = ps, data = mockstudy, subset = ps %in% 0:1)
summary(tb)
```

\normalsize

# tableby

Other features:

- `as.data.frame()`, `as.data.frame(summary())`

- Subset variables, change the order, delete variable: `[`, `head()`, `tail()`

- Sort by p-value: `sort()`

- Filter by p-value: `<`, `<=`, `>`, `>=`, etc.

- Merge two tables: `merge()`

- Custom p-values and user statistics

# modelsum

The basic table (modeling `alk.phos ~ arm` and `alk.phos ~ ps`)

\tiny

```{r results='asis'}
ms <- modelsum(alk.phos ~ arm + ps, data = mockstudy)
summary(ms)
```

\normalsize

# modelsum

Add common adjusters:

\tiny

```{r results='asis'}
ms <- modelsum(alk.phos ~ arm + ps, data = mockstudy, adjust = ~ sex + age)
summary(ms)
```

\normalsize

# modelsum

Don't show common adjusters:

\tiny

```{r results='asis'}
ms <- modelsum(alk.phos ~ arm + ps, data = mockstudy, adjust = ~ sex + age)
summary(ms, show.adjust = FALSE, show.intercept = FALSE)
```

\normalsize

# modelsum

Other options:

- Change model "family": Poisson, Binomial, Survival, Negative Binomial, Ordinal, Conditional Logistic, Relative Risk

- Change labels: `labels()<-`, `set_labels()`, `labelTranslations=`

- Change decimal places: `modelsum.control()`

- Change summary statistics: `modelsum.control()`

- `as.data.frame()`, `as.data.frame(summary())`

- Subset variables, change the order, delete variable: `[`, `head()`, `tail()`

- Merge two tables: `merge()`

# freqlist

\tiny

```{r results='asis'}
fl <- freqlist(~ sex + arm + ps, data = mockstudy)
summary(fl)
```

\normalsize

# freqlist

Sorted by frequency:

\tiny

```{r results='asis'}
fl <- freqlist(~ sex + arm + ps, data = mockstudy)
summary(sort(fl), dupLabels = TRUE)
```

\normalsize

# freqlist

Other options:

- `as.data.frame()`, `as.data.frame(summary())`

- Change labels: `labels()<-`, `set_labels()`, `labelTranslations=`

- Subset variables, change the order, delete variable: `[`, `head()`, `tail()`

- Merge two tables: `merge()`

# comparedf

\tiny

```{r}
mockstudy2 <- muck_up_mockstudy()
cdf <- comparedf(mockstudy, mockstudy2, by = "case")
cdf
# summary(cdf)
```

\normalsize


# comparedf

\tiny

```{r}
cdf <- comparedf(mockstudy, mockstudy2, by = "case", int.as.num = TRUE,
                 factor.as.char = TRUE, tol.vars = "case")
cdf
```

\normalsize


# comparedf

\tiny

```{r}
tail(diffs(cdf))
diffs(cdf, by.var = TRUE)

```

\normalsize

# write2

Three main functions: `write2word()`, `write2pdf()`, `write2html()`

Can use other output formats supported by R Markdown.

\tiny

```{r results='hide', message=FALSE}
write2pdf(list(
  tb,
  summary(lm(age ~ sex, data = mockstudy)),
  "\\newpage",
  "# My modelsum table",
  ms,
  code.chunk(1 + 1)
), file = "test.pdf")
```

\normalsize


# Resources


Docs: https://mayoverse.github.io/arsenal/

Issues: https://github.com/mayoverse/arsenal/issues/

This presentation: https://github.com/eheinzen/2021_arsenal_RMedicine/tree/Recording

Connect with us on Github: `@eheinzen`, `@bethatkinson`, `@sinnweja`, https://github.com/mayoverse/







